---
title: "01 Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01 Get started}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# 1) Installation and Technical Requirements

## Introduction

In terms of natural language processing (NLP), state-of-the-art approaches rely on neural nets
with multiple layers and consist of a huge number of parameters making
them computationally demanding. With specialized libraries such as
*PyTorch*, graphical processing units (gpu)
can help to speed up computations significantly. However, many of these
specialized libraries for machine learning are written in python.

The R package *Artificial Intelligence for Education (aifeducation)*
aims to provide educators, educational researchers, and social
researchers with a convincing interface to these state-of-the-art models
for natural language processing and tries to address the special needs
and challenges of the educational and social sciences. The package
currently supports the application of Artificial Intelligence (AI) for
tasks such as text embedding and classification.

Since state-of-the-art approaches in natural language processing rely on
large models compared to classical statistical methods (e.g., latent
class analysis, structural equation modeling) and are based largely on
python, some additional installation steps are necessary.

If you would like to train and to develop your own models and AIs, a
compatible graphic device is necessary. Even a low performing graphic
device can speed up computations significantly. If you prefer to use
pre-trained models however, this is **not** necessary. In this case, a
'normal' office computer without a graphic device should be sufficient
in most cases. To get ready for using the package, two steps are
necessary.

## Step 1 - Install the R package

First you need to install the package. This can be done by:

```{r, include = TRUE, eval=FALSE}
install.packages("aifeducation")
```

With this command, `aifeducation` is installed on your machine.

## Step 2 - Install Python and optional R packages

Since natural language processing with neural nets is based on models
which are computationally intensive, *PyTorch* is used within this
package together with some other specialized python libraries.

> Before you start installing python or updating python packages, please
> ensure that you have the newest version of the *R* package 'reticulate' installed.

The most straightforward method for getting started is to call the
function `install_aifeducation` as follows:

```{r, include = TRUE, eval=FALSE}
aifeducation::install_aifeducation(
  install_aifeducation_studio = TRUE,
  python_version = "3.12",
  cuda_version = "12.4",
  use_conda = FALSE
)
```

This function will install python, and all relevant python
libraries into a virtual environment called "aifeducation".

In addition, we recommend to set `install_aifeducation_studio=TRUE`
since this will install the optional *R* packages necessary to use *AI
for Education - Studio*. *AI for Education - Studio* is a graphical user
interface for using this package. We recommend to use it for
everyone who is unfamiliar with *R* or machine learning. If you do not
want use studio, you can set the argument to `FALSE`. In this case
you have to use the package with *R* syntax.

If you have a suitable machine and would like to use a graphic card for
computations you need to install some further software. You can find
further information here: <https://pytorch.org/get-started/locally/>

You can check if python is working by using the function
`reticulate::py_available()`. This should return `TRUE`.

```{r, include = TRUE, eval=FALSE}
reticulate::py_available(initialize = TRUE)
```

You can check if all necessary python packages are successfully
installed by calling the function `check_aif_py_modules`

```{r, include = TRUE, eval=FALSE}
aifeducation::check_aif_py_modules()
```

Now everything is ready to use the package.



## Step 3 - Install Protocol Buffer Compiler (Optional)

Next you *can* install Protocol Buffer Compiler if it is not already on your system.
This compiler is a prerequisite for two python packages ('protbuf' and 'sentencepiece') 
and some open source models to work. These packages are also necessary to update
some older open source models to newer version of the library 'transformers'. 
You can find installation instructions on [https://protobuf.dev/installation/].

As an alternative you can call `aifeducation::install_protocol_buffer_compiler()`
for an automatic installation. 

After a successful installation of Protocol Buffer Compiler you can install
the two python packages by calling `reticulate::py_install(c("protobuf","sentencepiece"))`.
Please restart the session.

> Please note that this installation step is only necessary if you plan to
> work with base models that use a tokenizer based on 'sentencepiece'. You can found
> a table with all transformers models and their prerequisites in 
> [A02 Supported Base Models](https://fberding.github.io/aifeducation/articles/a02_base_model_types.html)

> Please note that all base models created with this package do **not** need
> these packages.

In order to be fully flexible we recommend  to install these packages and the compiler.

# 2) Starting a new session

The most convenient way to work with the package is to use *AI for
Education - Studio* which you can start by calling
`aifeducation::start_aifeducation_studio()`.

```{r, include = TRUE, eval=FALSE}
aifeducation::start_aifeducation_studio()
```

In case you do not want to use the graphical user interface, you have to
prepare your *R* sessions. The easiest way is to call the following function
directly after starting a *R* session.

```{r, include = TRUE, eval=FALSE}
library(aifeducation)
aifeducation::prepare_session()
```

> **Note:** Please remember: Every time you start a new session in *R,*
> you have to call the funcion `aifeducation::prepare_session`. This is not
> necessary if you use *AI for Education - Studio*.

# 3) Tutorials and Guides

A guide on how to use the graphical user interface can be found in
vignette [02 Using the graphical user interface Aifeducation -
Studio](https://fberding.github.io/aifeducation/articles/gui_aife_studio.html).

A short introduction into the package with *R* syntax can be found in vignette [03 Using R
syntax](https://fberding.github.io/aifeducation/articles/classification_tasks.html).

Documenting and sharing your work is described in vignette [04 Sharing
and using trained AI/models](https://fberding.github.io/aifeducation/articles/sharing_and_publishing.html).

# 4) Update *aifeducation*

In case you already use *aifeducation* and you want to update to a newer
version of this package, it is recommended to update the used python
libraries as well. The easiest way is to call the following function:

```{r, include = TRUE, eval=FALSE}
aifeducation::update_aifeducation(
  update_aifeducation_studio = TRUE,
  env_type = "auto",
  cuda_version = "12.4",
  envname = "aifeducation"
)
```

> Starting with version 1.1.2 the python package 'calflops' is necessary. If you
> used aifeducation with version 1.1.1 or lower please 
> install this package manually with `reticulate::py_install("calflops")` or call
> `aifeducation::update_aifeducation`.

> Starting with version 1.1.6 the python packages 'protobuf' and 'sentencepiece ' are optional packages. If you
> used aifeducation with version 1.1.5 or lower please 
> install first Protocol Buffer Compiler via `install_protocol_buffer_compiler` or
> (https://protobuf.dev/installation/)[https://protobuf.dev/installation/].
> After that install the packages manually with `reticulate::py_install(c("protobuf","sentencepiece"))`.

# 5) Recommended Version of Python Packages

Using the function above will install the newest and recommended version of the 
core python packages in order to use this package. In the case you would like to
install the packages manually we recommend the following versions.

```{r, include = TRUE, eval=TRUE}
t(aifeducation::get_recommended_py_versions())
```

We have tested the package with these versions ensuring that *aifeducation* works correctly.
